---
title: "Home Values After Transit Line"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(scipen=999)
library(tidyverse)
```

## Measuring the Impact of Transit on Housing Prices

Local officials have claimed that a new light rail line has increased property values in a local neighborhood.

Before the transit line opened, homes in that neighborhood sold for an average of **$285,000**, with a standard deviation of **$32,000**.

Since the extension opened, your news team collected sales data from 14 recent home sales in the neighborhood:
$297,000, $305,000, $292,000, $318,000, $290,000, $311,000, $299,000, $325,000, $302,000, $288,000, $307,000, $315,000, $320,000, $294,000

As a journalist, you need to determine: **Is there statistical evidence that the transit line has actually increased home values? And if so, by how much?**

Follow along, running the code when directed. If you have questions, ask either in class or put them just below the section you have a question about.

### Task 1: Organize your data

```{r}
# Known information about home values before the transit line
prior_mean <- 285000  # average home value
prior_sd <- 32000     # standard deviation

# Home sale prices after the transit line opened (14 homes)
new_prices <- c(297000, 305000, 292000, 318000, 290000, 311000, 299000, 
                325000, 302000, 288000, 307000, 315000, 320000, 294000)

# Create a journalist-friendly dataset
home_data <- tibble(
  sale_id = 1:14,
  sale_price = new_prices
)

# View the data
home_data
```

### Task 2: Calculate key statistics (1 point)

Replace the REPLACE_ME values to calculate the mean, standard deviation and count of the new home prices.

```{r}
new_stats <- home_data |> 
  summarise(
    mean = mean(sale_price),
    sd = sd(sale_price),
    n = n()
  )

new_stats
```

**Answer this question:** How does the new average compare to the prior average of $285,000? How much higher is it in dollar terms? Does that seem like a big difference to you? What else would you want to know?

The new mean is higher by about $20,000, which seems like a big difference to me. The standard deviation has also decreased into about a third of its original value. 

### Task 3: Visualize the distribution (1 point)

Let's make a histogram of the new home prices with a reference line showing the old average. This helps us see the spread of the data and where the prior average falls relative to the new sales.

```{r}
# Function to format currency values
dollar_format <- function(x) {
  paste0("$", format(x/1000, big.mark=","), "K")
}

ggplot(home_data, aes(x = sale_price)) +
  geom_histogram(binwidth = 5000, fill = "seagreen", color = "white", alpha = 0.8) +
  geom_vline(xintercept = prior_mean, color = "darkred", size = 1.2, linetype = "dashed") +
  geom_vline(xintercept = new_stats$mean, color = "darkblue", size = 1.2) +
  annotate("text", x = prior_mean - 1000, y = 3.5,
           label = "Previous Avg ($285K)", hjust = 1, fontface = "bold", color = "darkred") +
  annotate("text", x = new_stats$mean + 1000, y = 3,
           label = paste0("New Avg (", dollar_format(new_stats$mean), ")"), 
           hjust = 0, fontface = "bold", color = "darkblue") +
  scale_x_continuous(labels = dollar_format) +
  labs(
    title = "Home Sales After Transit Line Extension",
    subtitle = "Dashed line shows previous neighborhood average of $285,000",
    x = "Sale Price",
    y = "Number of Sales",
    caption = "Source: County Property Records"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10)
  )
```

**Answer this question:** Look at where the old average (red dashed line) falls relative to the new sales. How many of the 14 homes sold for more than the old average? What does the histogram tell you about the spread of the new prices?

All of the 14 homes sold for higher than the old average, and the histogram tells us that the spread is relatively condensed for the new prices.

### Task 4: Set up and perform a hypothesis test (2 points)

This is where we formally test the officials' claim. Fill in the REPLACE_ME values for the hypotheses and the t-test.

**Hypotheses:**

- **H₀ (Null):** The average housing prices have not changed.
- **H₁ (Alternative):** The average housing prices have increased.

```{r}
# Set the significance level
alpha <- 0.05

# Perform a one-sample t-test
# Since we want to know if home values INCREASED, we use alternative = "greater"
t_test_result <- t.test(
  home_data$sale_price,
  mu = 285000,
  alternative = "greater"
)

t_test_result
```

**Answer this question:** Look at the p-value. Is it less than our alpha of 0.05? What does this mean for our null hypothesis — can we reject it? What can we now say about the officials' claim?

The p-value is much less than 0.05, which means we can reject the null hypothesis because our data is scientifically significant. It is a lot harder to confirm that property value has increased, but it would seem that it is the case. We should look at the significance of the change using Cohen's D. 

### Task 5: Interpret the results with a confidence interval (2 points)

Remember confidence intervals from last week? The t-test gives us one too. Let's pull together the key statistics for your story.

```{r}
# Extract the p-value
p_value <- t_test_result$p.value

# Calculate the 95% confidence interval (two-sided for the full range)
ci <- t.test(home_data$sale_price)$conf.int

# Key statistics for your story
story_stats <- tibble(
  `Previous average` = dollar_format(prior_mean),
  `New average` = dollar_format(mean(new_prices)),
  `Increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent change` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%"),
  `p-value` = p_value,
  `95% CI lower` = dollar_format(ci[1]),
  `95% CI upper` = dollar_format(ci[2]),
  `Significant?` = if_else(p_value < alpha, "Yes", "No")
)

story_stats
```

**Answer this question:** The 95% confidence interval tells us the range where the true average home price likely falls. Does that range include the old average of $285,000? What does that tell you?

The range in the 95% confidence interval is higher than the old average of $285,000, meaning we are 95% confident that the true mean is higher than the previous average. 

### Task 6: Calculate the effect size (3 points)

We've established that the increase is statistically significant — it's real, not due to chance. But how big is it, really? We'll calculate Cohen's d to find out. The formula is:

**d = (new mean - prior mean) / prior standard deviation**

Replace the REPLACE_ME values.

```{r}
effect_size <- (mean(new_prices) - prior_mean) / prior_sd

effect_size_label <- case_when(
  abs(effect_size) < 0.2 ~ "Negligible",
  abs(effect_size) < 0.5 ~ "Small",
  abs(effect_size) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `Cohen's d` = round(effect_size, 3),
  `Interpretation` = effect_size_label,
  `Dollar increase` = dollar_format(mean(new_prices) - prior_mean),
  `Percent increase` = paste0(round(((mean(new_prices) - prior_mean) / prior_mean) * 100, 1), "%")
)
```

**Answer this question:** The effect size is "medium." In dollar terms, the average increase is about $19,500. But remember that the standard deviation of home prices in this neighborhood was $32,000 *before* the transit line — meaning prices already varied by tens of thousands of dollars from sale to sale. Given that context, does a $19,500 increase seem large or modest? How does the effect size help you think about this differently than just looking at the dollar amount?

Given that the old standard deviation was a little less than double the amount that the average increased, this change seems a lot more modest, especially given our sample size of only 14. The effect size also seems to suggest that this is a modest change, which makes sense. 

### Task 7: What's the story? (6 points)

You now have all the statistical evidence. But writing the story requires more than numbers — it requires judgment.

**Part A:** The local officials are going to claim the transit line increased home values. Based on your analysis, is that claim supported? Write 2-3 sentences for the top of your story that accurately reflect both the statistical significance AND the effect size. Don't just echo the officials' talking points — interrogate them.

Local officials claim the new transit line is increasing home values, but the change seems modest. Since the extension opened, the average house cost seems to have increased by at least $12,000. 

**Part B:** A $19,500 increase is real and statistically significant, but what *else* could explain rising home prices in this neighborhood? List at least two alternative explanations that a good reporter would want to investigate before attributing the increase entirely to the transit line.

You would definitely want to look at the way the price of housing is rising with inflation in general. You would also want to look at other recent additions to the neighborhood or changes in local policy that may be confounding variables in the price increase. 


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
