```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

## Does Poverty Predict Math Scores? Testing a School Board Member's Claim

### Introduction

A school board member in Montgomery County, Pennsylvania has claimed at a public meeting that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests. That's a testable claim, and as data journalists, our job is to test it.

We have school-level data from Montgomery County, Pa. that includes each school's FARMS percentage (the share of students eligible for Free and Reduced-price Meal System) and the percentage of students who met or exceeded math expectations on state tests. We'll use linear regression to evaluate whether the school board member's claim holds up, and we'll learn something important about what happens when you trust a model before examining your data.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_pa.csv")
```

### Task 1: Examine the Data Before You Plot Anything

Before we visualize anything, let's look at what we're working with.

```{r}
summary(schools$`FARMS Pct`)

schools |> arrange(desc(`FARMS Pct`)) |> head(10)

schools |> arrange(`FARMS Pct`) |> head(10)
```

Look at the distribution of FARMS percentages. Do any values look unusual or suspicious? What might it mean if a school has a FARMS percentage of exactly 100? Write two or three sentences describing what you see and whether anything concerns you.

ANSWER HERE


### Task 2: Create a Scatterplot and Describe the Results

Now let's visualize the relationship between FARMS percentage and math scores for all schools.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County PA")
```

Describe what you see. Is there a clear trend? Does the regression line look like it represents the data well, or does it seem to be pulled in an odd direction? Connect what you observed in Task 1 to what you see here.

ANSWER HERE


### Task 3: Run the Regression Model

Let's build the linear regression model using all schools.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

There are three things to examine: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct.

The **p-value** tells you whether the relationship is statistically significant. The **Adjusted R-squared** tells you how much of the variation in math scores is explained by FARMS percentage — it's already a percentage, so a value of 0.025 means 2.5%, not 25%. The **coefficient** on FARMS Pct tells you the direction and size of the relationship: positive means higher FARMS predicts higher math scores; negative means higher FARMS predicts lower scores.

Report all three values and describe what they tell you. Does this model support the school board member's claim? Is this even a model worth using?

ANSWER HERE


### Task 4: What's Driving the Problem? Filter and Rerun.

In Task 1, you likely spotted schools with a FARMS percentage of 100. Those values are suspicious — in most districts, having *every* student on free/reduced lunch is unusual and may indicate a reporting artifact or data error. Let's see what happens when we exclude them.

```{r}
schools_filtered <- schools |> filter(`FARMS Pct` < 100)

model_filtered <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools_filtered)
summary(model_filtered)
```

Now compare the two models side by side:

```{r}
cat("Full dataset model:\n")
cat("  Coefficient:", coef(model)[2], "\n")
cat("  p-value:", summary(model)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model)$adj.r.squared, "\n\n")

cat("Filtered dataset model (FARMS < 100):\n")
cat("  Coefficient:", coef(model_filtered)[2], "\n")
cat("  p-value:", summary(model_filtered)$coefficients[2,4], "\n")
cat("  Adjusted R-squared:", summary(model_filtered)$adj.r.squared, "\n")
```

How did removing those schools change the model? Did the coefficient change direction? Did the p-value or R-squared improve meaningfully? What does this tell you about the influence of suspicious data values on a regression model?

ANSWER HERE


### Task 5: Even the Better Model Has Problems

Even after filtering, the Adjusted R-squared remains very low. This means FARMS percentage — even when the data is clean — explains very little of the variation in math scores across these schools.

Look at the filtered scatterplot:

```{r}
schools_filtered |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage (Excluding 100% FARMS Schools)")
```

With a low R-squared, the residuals from this model aren't very meaningful as labels of "over-performing" or "under-performing" — the model doesn't know enough to make that judgment reliably. But let's calculate them anyway to see the spread:

```{r}
schools_filtered <- schools_filtered |>
  mutate(
    predicted = predict(model_filtered),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools_filtered |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

What are the largest residuals? Given the model's weak R-squared, how much confidence would you have in labeling these schools as over- or under-performers? Why?

ANSWER HERE


### Task 6: Write the Story Pitch

You've now run two models, examined the data quality and seen that even a cleaned-up model explains very little of the variation in math scores. Return to the school board member's claim: that schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch (3-5 sentences) that addresses the following: Does the data support or contradict the claim? What data quality issues did you find? What would you need - different data, a different model, additional reporting - before you could responsibly publish a story about this claim? Who would you call first?

ANSWER HERE
