---
title: "HW10: Significance Testing & Effect Size"
name: Vada Mercer
---

You previously calculated the mean, margin of error and confidence interval for monthly car thefts in Prince George's County. Your editor now wants to know: it _looks_ like car thefts are decreasing lately, but is that change real, and how big is it? In this assignment, you'll use significance testing and effect size to answer those questions. Load the tidyverse and the data to get started.

### Task 1: Load the tidyverse and the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# LOAD THE TIDYVERSE
```

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

```

### Task 2: Split the data and test for significance (4 points)

We'll split the monthly data into two halves, and then ask whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05. First, fill in the hypotheses, then replace the REPLACE_ME values in the code.

**Null Hypothesis:** The average amount of car thefts has not changed.

**Alternative Hypothesis:** The average amount of car thefts is decreasing.

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |>
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

```

**Answer this question:** What is the p-value? Is it less than 0.05? Can we reject the null hypothesis? What does this tell your editor?

The p-value is below 0.05, which means we can reject the null hypothesis that the average amount of car thefts has not changed. We are able to tell our editor that it is highly likely the average amount of car thefts is changing,

### Task 3: Visualize the comparison (4 points)

Let's see the two periods side by side with their confidence intervals. Replace the REPLACE_ME values.

```{r}
# Summarize by period
period_summary <- car_thefts |>
  group_by(period) |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = 32 - 1) * se_thefts,
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |>
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 5)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Answer this question:** Look at the error bars (confidence intervals) on the chart. Do they overlap? What does that tell you about whether the difference between the two periods is real?

The confidence intervals do not overlap at all, which helps support the theory that there is a genuine difference in average car thefts between these two periods of time, and that we can be almost certain of it.  

### Task 4: Calculate the effect size (5 points)

We've established that the decline is statistically significant. But how large is it? Let's calculate Cohen's d to measure the effect size. For a two-sample comparison, we use the pooled standard deviation:

**Pooled SD** = sqrt(((n1 - 1) × sd1² + (n2 - 1) × sd2²) / (n1 + n2 - 2))

**Cohen's d** = (mean1 - mean2) / pooled SD

Replace the REPLACE_ME values. You'll need to pull the values from `period_summary`.

```{r}
# Extract the values from period_summary
first_half_stats <- period_summary |> filter(period == "First Half")
second_half_stats <- period_summary |> filter(period == "Second Half")

# Calculate pooled standard deviation
pooled_sd <- sqrt(
  ((first_half_stats$months - 1) * first_half_stats$sd_thefts^2 +
   (second_half_stats$months - 1) * second_half_stats$sd_thefts^2) /
  (first_half_stats$months + second_half_stats$months - 2)
)

# Calculate Cohen's d
cohens_d <- (first_half_stats$mean_thefts - second_half_stats$mean_thefts) / pooled_sd

# Interpret it
effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `First half mean` = round(first_half_stats$mean_thefts, 1),
  `Second half mean` = round(second_half_stats$mean_thefts, 1),
  `Difference` = round(first_half_stats$mean_thefts - second_half_stats$mean_thefts, 1),
  `Cohen's d` = round(cohens_d, 3),
  `Interpretation` = effect_label
)
```

**Answer this question:** How would you describe this effect size? Compare it to the effect sizes we calculated in class (the response times activity had a small effect, the housing prices activity had a medium effect). What does a large effect size mean for the story you'd write about car thefts?

The effect size is large, and much larger than I've seen in previous exercises. This tells us that there is almost certainly a change in the true average amount of car thefts and that putting out a story stating car thefts have gone down in the area would be a compeletely reasonable thing. 

### Task 5: Putting it all together (7 points)

You now have three pieces of evidence: the confidence intervals from last week, the significance test from this assignment, and the effect size.

1. In Part 1, your editor asked "just tell me if car thefts are going up or down". Now you have significance testing. Write 2-3 sentences telling your editor the answer, using all three pieces of evidence (CI, p-value, and effect size).

Dear editor, 

I am fairly certain car thefts are going down in the area. When splitting the data into two periods, we found a much higher average car thefts value in the first period than in the more recent one, and we can guarantee with 95% certainty that the true average amount of car thefts has changed. 

2. Write a headline and a 2-3 sentence lede for a story about car thefts in Prince George's County. Unlike the response times story from class (where the effect was small and the headline needed to be cautious), this is a case where the data strongly supports a clear conclusion. Your writing should reflect that — but still note the limitations (correlation vs. causation, what else might explain the decline).

Headline: Car thefts down in Prince George's County since 2023

The average amount of car thefts was found to have decreased comparing data within the last 3 years. When split into two periods, the average amount of car thefts in the first period was found to be 438.9 a month, while the average amount in the most recent period was at 312.2 a month. 

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
