---
title: "Identifying Schools with Biggest Differences in Math Test Scores, Part 2"
author: "YOUR NAME"
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen = 999)
```

### Introduction

In the class activity, you tested a school board member's claim using data from Montgomery County, Pa. — and found a model that didn't hold up. Now you'll run the same analysis using data from Montgomery County, Md. Your job is to evaluate whether the Maryland data tells a different story, and what that means for how you'd report on the claim.

### Loading the Data

```{r}
schools <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/montgomery_md.csv")
```

### Task 1: Examine the Data First (2 points)

Before you plot anything, look at the data — you know why by now.

```{r}
summary(schools$`FARMS Pct`)
schools |> arrange(desc(`FARMS Pct`)) |> head(10)
```

Are there any data quality issues like the ones you found in the Pennsylvania data? How does the distribution of FARMS percentages compare between the two counties? Note anything that concerns you or gives you confidence in the data.

Since the highest FARMS percent is at 96, and since most schools with high amount of FARMS percentages still have a fair amount of students that do not qualify, I am a lot more confident in the zoning of these schools and the standards for FARMS being accurately filled.


### Task 2: Create a Scatterplot and Compare (5 points)

Replace the REPLACE_ME value with the correct variable name to create a scatterplot of FARMS percentage vs. math scores for Montgomery County, Md.

```{r}
schools |>
  ggplot(aes(x = `FARMS Pct`, y = `% Met or Exceeded Math Expectations`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "FARMS Percentage", y = "% Met or Exceeded Math Expectations",
       title = "Math Scores vs. FARMS Percentage, Montgomery County MD")
```

Describe what you see. Then compare this scatterplot specifically to the Pennsylvania one: what differences do you notice in the direction, strength, and spread of the relationship? The two counties share a name and are both suburban — what might explain why they look so different?

This scatterplot has a much more clear downward slope. Though still not completely perfect, there is an obvious trend in the data that shows the percent meeting or exceeding math expectations tends to decrease in schools when the amount of students qualifying for free and reduced meals increases. This data has a lot more variance in the FARMS pct area, there are schools with higher percentages of FARMS students that do not reach 100%.


### Task 3: Linear Regression Model (5 points)

Replace the REPLACE_ME value with the correct variable name to run a linear regression model predicting math scores from FARMS percentage.

```{r}
model <- lm(`% Met or Exceeded Math Expectations` ~ `FARMS Pct`, data = schools)
summary(model)
```

Report the three key statistics: the p-value, the Adjusted R-squared, and the coefficient for FARMS Pct. Then interpret them together — not separately. What does the combination of these three numbers tell you about the relationship between poverty and math scores in Montgomery County, Md.? Is this model worth using? How does it compare to the Pennsylvania model?

The p-value is 0.00000000000000022, telling us this data is extremely scientifically significant. The adjusted R-squared is 0.6734, telling us that 67.34% of the change in the percent meeting or exceeding math expectations is explained by the percent of students on FARMS. The coefficient is -0.59511, saying for every one percentage point extra in students on FARMS, the percent meeting or exceeding math standards decreases by 0.59511.



### Task 4: Identifying Schools with the Biggest Residuals (6 points)

Replace the REPLACE_ME values with the correct variable names to calculate predicted values, residuals, and absolute residuals for each school. Then arrange the schools in descending order of absolute residuals.

```{r}
schools <- schools |>
  mutate(
    predicted = predict(model),
    residual = `% Met or Exceeded Math Expectations` - predicted,
    abs_residual = abs(residual)
  ) |>
  arrange(desc(abs_residual))

head(schools |> select(1, `FARMS Pct`, `% Met or Exceeded Math Expectations`, predicted, residual), 10)
```

Identify the school with the largest positive residual (doing much better than the model predicts) and the school with the largest negative residual (doing much worse). For each school, find its zip code and look it up on [CensusReporter.org](https://censusreporter.org).

For each of the two schools, answer:
- What does the demographic and economic profile of the zip code look like?
- Does anything in that profile help explain why the school is over- or under-performing the model?
- What would your first reporting question be to the principal of that school?

The school with the largest positive residual is Fox Chapel Elementary, with a 29.99 residual. I would want to ask the principal what the school's approach to education is and their academic policies. Despite 56.7% of students qualifying for FARMS, it seems the poverty rate for children under 18 is about 21%. The poverty rate for senior citizens in this community is much lower at 9%, which suggests that there is a strong support system for struggling students. 

The school with the largest negative residual is Germantown Elementary, with a -28.08 residual. I would also want to ask the principal what the school's approach to education is and their academic policies, as well as their plans for getting students back on track with math. I notice that this district has a higher-than-average amount of foreign born people in their population at 37.1%, so I wonder if the transition between educational systems is causing some struggle for these students. I would definitely want to also ask the principal what their program is for bringing in students who are unfamiliar with the American education system. 


### Task 5: Evaluate the Model and Write the Pitch (7 points)

You've now run this analysis on two counties. Return to the school board member's claim: that elementary schools with fewer students on free and reduced-price meals do better on standardized math tests.

Write a short story pitch of 4-6 sentences that addresses all of the following: Does the Maryland data support or contradict the claim? How does it compare to what you found in Pennsylvania? What would you ask the school board member if you called them for comment? And what's the most interesting specific finding from your residuals analysis: which school would you want to visit first, and why?

The Maryland data does strongly support the claim that a school with more students on free and reduced meals is more likely to underperform in math expectations. The data in Pennsylvania, however, suggested that there was little correlation, suggesting that this correlation varies geographically and may be influenced by confounding factors. I wouldn't be opposed to reporting on Fox Chapel's performance outdoing the predicted value by about 30 points, I think this is an important detail to include and remind readers that things aren't always so black and white and that there are things being done to remedy the issue. I would want to speak to school board members about their opinions on the claim, providing them with evidence to back it up, and ask what they perceive the issue and the solution as. 
