---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE

The stories that will always make the news and that people will be out for will be homicide stories, as well as shootings. Carjacking and other assault statistics will also be notable.S One issue I have with this dataset is the "accident" classification, "accident" and "accident with impound." I want to know if the accidents with impounds are factored into the regular accident statistic. I have a similar issue with the "auto stolen & recovered," as well as "theft" and "theft from auto." I'm also confused what "theft from auto" entails and what makes it different from "auto stolen."

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> mutate(mean = (Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+Dec)/12)
#I feel like there was an easier way to calculate this than having to type all these out, but I'm not sure how that would be done
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

The three most reported crimes in Prince George's County were accidents, "auto stolen," and "theft from auto." There were an average of 481.33 accidents every month and 320.75 with impound. If these are exclusive to each category, that's an average of 802.08 accidents every month. There was an average of 399.167 stolen automobiles every month, with 40.583 being recovered stolen automobiles every month. If we assume these are exclusive, that would mean there was an average of 439.75 automobile thefts every month. Theft from automobiles, which I assume means taking items from a car, averaged at 391.917, while other general theft was at 284.083. Assuming these are exclusive, that means there was an average of 676 thefts every month. This data suggests automobile owners may be more likely to be victims of crime. Of course the question still remains which of these is exclusive, I don't think you could write a whole story (or at least confidently publish one) with just this table of information. 

ANSWER HERE


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate (june_change = ((Jun-mean)/mean)*100) |> mutate (august_change = ((Aug-mean)/mean)*100) |> mutate (july_change = ((Jul-mean)/mean)*100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

There were a few crimes that were consistently up all three months of summer, which were homicide, automobiles stolen and theft from autmobiles. Some crimes remained down all months of summer, these being breaking and entering (other) and theft. There were certain crimes that increased in some summer months but decreased in another, like how vehicle robbery was up 24.41% in June compared to the average, up 64.55% in July, then fell 15.72% below the average in August. To comfortable assert the significance of these figures, I'd want to know the standard deviation of these crimes, because right now it really seems like a mixed bag, and I don't notice too much of a trend. 

ANSWER HERE

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE (For this story, I had to make some assumptions about the data presented that I would usually want to clarify with an editor or with the source of information about. I also created a new dataset, as I thought offering totals instead of averages would be a better way to present this information)

Automobile related crime has been the most frequently reported crime this past year in Prince George's County.

With 9,625 reported accidents, 4,790 reported instances of stolen automobiles, and 4,703 reported instances of theft from an automobile this past year, automobile owners in Prince George's County are among the most affected residents of crime, highlighting growing concerns of vehicle safety across the county. 

```{r}
crime_totals <- pg_crime |> mutate (total=(Jan+Feb+Mar+Apr+May+Jun+Jul+Aug+Sep+Oct+Nov+Dec))
```



### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
