---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library('tidyverse')
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read.csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Since there are some people with a salary of 0, this may skew the mean if we look for an average salary among this data. I'm also curious why these people are reported as having a $0 salary, if that is the truth of their income and they remain on the employee list without working or if there was an error.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY))
wh_salaries |> summarize(median_salary = median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary = mean(SALARY))
wh_salaries |> summarize(median_salary = median(SALARY))
wh_salaries |> summarize(min_salary = min(SALARY))
wh_salaries |> summarize(max_salary = max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

I was surprised by the mean being higher than the median, because I would have thought the few 0 dollar salaries would drag it down compared to the other salaries in the hundred thousands. The mean was 109117.40, and the median was 101500. The maximum salary is interesting to me because it is more than double the amount of the median and mean salary, so whatever work this individual is doing is probably very important, or she is drawing on a lot of experience for it. She is making more than 20k more than the person below her, which is kind of strange considering every other salary seemed to be gradually building up. The minimum salaries of $0 confuse me, and I wonder if these people are making money in ways other than a salary

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

The question I get from this data is "who is Jacalynne B. Klopp? what is she doing as an "advisor" that is earning her more than double both the average and median salary? What experience is she drawing upon as an advisor, and why is that experience so expensive?" I'd start by doing some digging online and reach out to a spokesperson at the whitehouse for comment.
