---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library('tidyverse')
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(
  violent_rate_2019 = (violent2019 / pop2019)*100,
violent_rate_2020 = (violent2020 / pop2020)*100,
property_rate_2019 = (property2019 / pop2020)*100,
property_rate_2020 = (property2020 / pop2020)*100)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

Ocean City led with highest rate of property crime in 2019 and 2020, with 10.4 property crimes for every 100 people in 2019 and 12.13 for every 100 people in 2020. They also had the highest rate of violent crime in 2020, with a rate of 2.40 violent crimes per 100 people. Baltimore had the highest violent crime rate in 2019 of 1.83 violent crimes for every 100 people. Ocean City was second behind Baltimore in 2019 for violent crime with a rate of 1.31 violent crimes for every 100 people. Baltimore was second behind Ocean City for violent crime in 2020 at 1.60 for every 100 people. Those two cities are definitely noteworthy, and looking at the crime there is important. It surprised me that the rate of property crime in Ocean City rose in 2020, due to COVID I would have expected it to decrease. 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_change = ((violent_rate_2020 - violent_rate_2019)/(violent_rate_2019))*100, property_change = ((property_rate_2020 - property_rate_2019)/(property_rate_2019))*100)
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The most interesting news stories are usually the ones based on things that are unusual. With very low numbers, like violent crime going from 1 to 0, it will look like a huge 100% change, so just because the percent change looks big may not mean the change is that significant. There are some things that stand out as unusual to me, like violent crime in Edmonson rising almost 900%, however if you look at the true numbers it had gone from 3 violent crimes in 2019 to 12 in 2020, and the rate for violent crime was still under 1 per every 100 people in both instances. I think the most interestiing increase shown in this data is the 318.18% increase in property crime seen in Crisfield, rising from 0.43 property crimes for every 100 people to 1.81 for every 100 people. The raw numbers show there were 11 property crimes in 2019 and 46 property crimes in 2020. 

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

Despite lockdown from the COVID-19 pandemic, property crime rates have jumped more than 300% this past year in Edmonson, MD. 

Rates rose from 0.43 property crimes for every 100 people to 1.81 property crimes for every 100 people in the city.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Percentiles
Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.
Do the results change your best story idea from the previous assignment? Why or why not?

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentiles = ntile(violent_rate_2020, 100))
md_cities_percentiles |> summarize(
  iqr = IQR(violent_rate_2020)
  
)
### iqr = 0.4198895
```

10 cities are above the 90th percentile, meaning 10 cities had a higher violent crime rate than 90% of other Maryland cities.
The interquartile range is 0.4198895, meaning the middle 50% of violent crime rates are spread over a range of about 0.42. Seeing as our highest value is only about 1.31, this spread means there is a moderate amount of variation in the data. I still think I would keep my story about the property crime rate rising. 
