---
title: "HW7b - Probability and the Normal Distribution"
---

In this assignment, you'll build on your z-score skills to answer probability questions using the normal distribution. You'll return to the White House salaries data and use it to answer the kinds of "what are the chances?" questions that come up in journalism all the time. This sets the stage for sampling, which we'll cover next.

### Task 1: Load the tidyverse and the data (1 point)

Load the tidyverse and the White House salaries data from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv into a dataframe called `wh_salaries`

```{r}
library(tidyverse)
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```

### Task 2: Filter and summarize (2 points)

You may remember from HW3 that some White House staffers have a salary of $0. Let's filter those out and focus on paid employees, saving the result to a new dataframe called `wh_paid`. You need to calculate the mean, median and standard deviation of the salary column using `wh_paid`:

```{r}
wh_paid <- wh_salaries |>
  filter(SALARY > 0)
wh_paid |> 
  summarize(
  mean_sal = mean(SALARY),
        median_sal = median(SALARY),
         sd_sal = sd(SALARY))


```

Now create a histogram of the paid salaries with a vertical line for the mean and two more for ±1 standard deviation, similar to what you did in HW6 with Spotify data.

```{r}
wh_paid |> 
  ggplot() + 
  geom_histogram(aes(x = SALARY), binwidth = 10000)+
  geom_vline(aes(xintercept = mean(SALARY)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) - sd(SALARY)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(SALARY) + sd(SALARY)), color = "blue", linetype = "dashed", size = 1)
```

**Describe the shape of the distribution. Is it roughly normal? Skewed? Where do most salaries cluster?**

This shape does not appear to be normal, the majority of salaries seem to lie one standard deviation below the mean amount of salaries around the 65,000 point. The graph seems to be skewed to the left.

### Task 3: The Empirical Rule (3 points)

The empirical rule (also called the 68-95-99.7 rule) says that for a normal distribution:
- About 68% of data falls within 1 standard deviation of the mean
- About 95% falls within 2 standard deviations
- About 99.7% falls within 3 standard deviations

Let's test this. We'll calculate the salary ranges for 1, 2, and 3 standard deviations from the mean, then count how many employees actually fall in each range.

```{r}
mean_sal <- mean(wh_paid$SALARY)
sd_sal <- sd(wh_paid$SALARY)
total_n <- nrow(wh_paid)

empirical <- tibble(
  range = c("Within 1 SD", "Within 2 SD", "Within 3 SD"),
  lower = c(mean_sal - sd_sal, mean_sal - 2 * sd_sal, mean_sal - 3 * sd_sal),
  upper = c(mean_sal + sd_sal, mean_sal + 2 * sd_sal, mean_sal + 3 * sd_sal),
  expected_pct = c(68, 95, 99.7),
  actual_count = c(
    sum(wh_paid$SALARY >= (mean_sal - sd_sal) & wh_paid$SALARY <= (mean_sal + sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 2 * sd_sal) & wh_paid$SALARY <= (mean_sal + 2 * sd_sal)),
    sum(wh_paid$SALARY >= (mean_sal - 3 * sd_sal) & wh_paid$SALARY <= (mean_sal + 3 * sd_sal))
  ),
  actual_pct = round((actual_count / total_n) * 100, 1)
)

empirical
```

**How closely does the data follow the empirical rule? What does any deviation tell you about the distribution?**

As opposed to 95%, 100% of the data falls within 3 standard deviations, which makes sense because the lowest end is an impossible negative figure, and the upper end is just shy above the top salary. The actual percent of data within the first and second standard deviations are lower than expected, which suggests our data may not be normal, however the actual percents are only off from the expected percents by very few percentage points (about 2 and about 4 for the first and second standard deviations) This suggests our data could be close to normal, though it seems more people are outside of the 2 standard deviations than expected. (8.8% are outside of 2 standard deviations)

### Task 4: From z-scores to probability with pnorm() (4 points)

Remember the car thefts z-score activity from class? Well, I want to return to that briefly, introduce a new tool called pnorm() and then apply it here. So open that CarTheftZScores notebook and, at the bottom, copy and paste in the contents here: https://gist.githubusercontent.com/dwillis/25102889f576b1707576b85e0be0c353/raw/1ca2be9636497cb5293ff7b1f4e6ed8e27a9e049/car_theft_pnorm.rmd

Re-run all of the blocks in CarTheftZScores notebook, including the code you just pasted in. What's happening there is you are using a function called `pnorm()` to find the probability of a value being less than or equal to a certain z-score. The basic formula to find that probability is:

```
1 - pnorm(z_score)
```

Where `z_score` is one row's z-score. Now apply the same approach to salaries.

Your editor asks: **"What's the probability that a randomly selected White House employee makes more than $150,000?"**

To answer this:
1. Calculate the z-score for $150,000
2. Use `pnorm()` to find the probability. `pnorm()` gives you the area under the normal curve to the LEFT of a z-score — that is, the probability of being BELOW that value.
3. Subtract from 1 to get the probability of being ABOVE $150,000
4. Check your answer against the actual data

```{r}

salary <- 150000
z_score <- (salary - mean_sal) / sd_sal

prob_less <- pnorm(z_score)
prob_more <- 1 - prob_less
prob_more

# Check against actual data
actual_prob <- sum(wh_paid$SALARY > 150000) / total_n
actual_prob
```

**Write a sentence describing this probability in plain English. How does the normal approximation (prob_less) compare to the actual percentage?**

If we were to choose a white house worker at random, we would expect there to be a 16.9% chance that the worker we chose has a salary over $150,000, however, we instead found that there is a 18.2% chance that the worker we chose at random has a salary over $150,000.


### Task 5: Probability between two values (3 points)

Now your editor asks: **"What percentage of White House employees make between $80,000 and $120,000?"**

You'll need to calculate z-scores for those two figures, use `pnorm()` for each, and subtract to find the probability between them. Update the REPLACE_ME figures with the appropriate numbers.

```{r}

z_lower <- (80000 - mean_sal) / sd_sal
z_upper <- (120000 - mean_sal) / sd_sal

prob_between <- pnorm(z_upper) - pnorm(z_lower)
prob_between

# Verify with actual data
actual_between <- sum(wh_paid$SALARY >= 80000 & wh_paid$SALARY <= 120000) / total_n
actual_between

```

**Interpret this result. What does this "middle range" tell us about the White House salary structure?**

While we expected to find 36.6% of white house workers salaries falling within $80,000 to $120,000, the actual percent was less-- 32.6% of workers have a salary within this range. This further backs the claim that the white house salaries may be skewed and is not normally distributed. 

### Task 6: Expected vs. observed (3 points)

Let's compare what the normal distribution *predicts* to what we actually *see*. This idea — comparing expected to observed — is one you'll use again later in the course.

Create these salary bins and calculate the expected count (from the normal distribution) and the observed count (from the actual data) for each:

```{r}
bins <- tibble(
  category = c("Below $60k", "$60k-$90k", "$90k-$120k", "$120k-$150k", "Above $150k"),
  lower = c(-Inf, 60000, 90000, 120000, 150000),
  upper = c(60000, 90000, 120000, 150000, Inf)
) |>
  mutate(
    expected_pct = pnorm(upper, mean = mean_sal, sd = sd_sal) - pnorm(lower, mean = mean_sal, sd = sd_sal),
    expected_count = round(expected_pct * total_n, 1),
    observed_count = c(
      sum(wh_paid$SALARY < 60000),
      sum(wh_paid$SALARY >= 60000 & wh_paid$SALARY < 90000),
      sum(wh_paid$SALARY >= 90000 & wh_paid$SALARY < 120000),
      sum(wh_paid$SALARY >= 120000 & wh_paid$SALARY < 150000),
      sum(wh_paid$SALARY >= 150000)
    ),
    difference = observed_count - expected_count
  )

bins |> select(category, expected_count, observed_count, difference)
```

**Where does reality differ most from what the normal distribution predicts? What might explain those differences?**

While there are about 40 people expected to have salaries below $60k, we find only 2 individuals have salaries below this. That makes sense, if you're working at the white house you probably deserve a salary higher than this, and the white house should definitely be able to pay it. We expect to find 78 individuals with salaries between $60k-$90k, but we find almost double-- 151 people with a salary in this range. It kind of makes sense, these are probably "lower value" jobs with a lower salary, but they get a higher quantity of workers because more work is needed for something like that. There are 14 more people than expected with a salary above $150k, at 81, which may be the number of individuals with "high importance" jobs that "deserve" a higher salary. 

### Task 7: Story pitch (4 points)

Based on your probability analysis, write a headline and lede for a story about White House salaries. Your lede should incorporate at least one probability or percentage you calculated.

Then write 2-3 sentences describing how you would use probability thinking to advance this story — what could you compare these probabilities to, and what additional analysis would help?

Headline: White House Salaries Cluster Below 6 Figures

While the average white house salary in 2025 was $111,321, the most frequently observed salary lays between $60,000 to $90,000. 

I would go on in this story to clarify where "the most frequently observed salary" statistic comes from, that we broke salaries down into ranges of $30k. I would also like to include that the frequency is almost double what is expected if salaries were normally distributed, though I do wonder if salaries are even expected to be normally distributed. I suppose to know that I would look at past white house salaries to see what looks to be the norm, though if I go too far back I may need to convert the salaries factoring in inflation. It could add some contrast if I wanted to include that there was also a not-insignificant number of staffers earning over $150k, this could add some context and contrast to the lower ended salaries. 



### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
