---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library('tidyverse')
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mnms_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset

```{r}
set.seed(8)
sample_5 <- mnms_data |> sample_n(5)
```


### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms
```{r}
proportion_5 <- mean(sample_5$black)
### 78.6
sd_5 <- sd(sample_5$black)
### 18.93
```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4
```{r}
proportion_mnms <- mean(mnms_data$black)
### 73.28
sd_mnms <- sd(mnms_data$black)
### 15.75
```


### Task 6: Compare the sample and population statistics

The mean value for black in the sample proportion was about 5 mnms short of the true proportion of black mnms, while the standard deviation was about 3 mnms larger than the true standard deviation for the population of black mnms. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This tells us that we can get kind of close, in this case about a third of a standard deviation away, to finding the true population's proportion when we sample, but taking a small sample will lead to higher standard deviations and thus more varied results. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
