---
title: "final_project"
output: html_document
---

### QUESTIONS

re: linear model
The slope is 0.6734, which is a positive number indicating some level of change. I was wondering if this is a significant slope in terms of my data, given that my x values reach such a high number I was thinking a decimal point could be insignificant, however I do remember there is a value we find that will be between -1 and 1, and if this is that value then I would assume it is significant. 



### TRIALS 
   Goals: Look at some graph options, do significance tests for the collective elections, maybe look to see if it has increased over time. 
   
# Significance test



### TESTS

The most obvious test I want to run is a significance test on the differences between the average difference of partisan votes for elections, with a null-hypothesis that the mean value for democratic and republican is 0. I am pretty confident this will prove conclusive, so I'm wondering how far in each direction I can disprove null-hypotheses. (i.e. is it an even further mean than if the true difference were 3 percentage points? something like that)

I would be interested in seeing if there is a correlation between time and partisan difference, so running a test for correlation would be good too. I'm also interested in seeing if general partisan divide, or the gap between democratic and republican votes in these counties, has increased over time. I would imagine I would have to mutate the chart a few times to find this. I'm honestly not entirely sure how I could run this test, but I have a vision for it that I would like to try out. What I'd imagine I have to do is mutate for each election to find the difference is ex: 2000_D - 2000_R = 2000_CtDif. And that would become a new column, and then each election would have one column. Then I would take the mean from each of the CtDifs and plot it and try to find the correlation coefficient. 
###INFO

I found the largest college from every state and looked at the voting results from the past 7 elections in the counties those colleges are located in. I removed Alaska from the dataset because Alaska does not have counties, and the population is very small and unique. 

https://www.businessinsider.com/map-university-college-most-students-each-state-2014-3 This is the source I used to find the largest schools, I am just realizing it is a little old but I feel like since 2014 is about the middle of 2000-2024 that it is okay. 

The Cook PVI scores would not load. I did not understand how to open the file with the per-county voting information, so my Excel sheet is entirely hand-entered. I have not finished entering my data, sorry. I recorded the proportions of Democratic and Republican popular votes from each election. I added the modifier  "university-- yes or no" because I figured it could help me filter information in R. Frankly, I think most of the computing is better done in Excel. I would like to find a way to take the difference of the proportions, I think those figures will be the key to analyzing this trend. I know how to do this in Excel, and may end up just modifying the data in Excel, but I understand if you would prefer I use R to do that. 

I also think that some graphics will be really important for my final, something showing the difference in proportion for college counties and their states. I am REALLY going to need help with that, I think ggplot is the most confusing R function. So I am looking for a lot of advice on this project, but as I enter data I am noticing a trend.

I have been noticing that a lot of these counties are voting much more Democratically than their states, some by almost 20 percentage points. I want to run a few tests to see the significance of this difference, I suppose the null hypothesis in those tests would be that the difference between college counties and their states' voting proportion would be 0. 

Since my sheet is unfinished, I didn't want to load it in here, but I will provide a link to the WIP.

This link gives you the ability to view, if you would prefer I send a link to edit, let me know!
https://1drv.ms/x/c/932aad8149925c8d/IQAwUqCIIcLCTqgzqPMuakQwAVIOPvJlHRUdpY7f-PNvduI?e=Ya5weg 



```{r}
library(tidyverse)
```

```{r} 
election_data <- read_csv("CollegeCountyParties.csv")
```

```{r}
election_data |> summarize(mean_2000_d = mean(D_2000))
```
### tidying the data using professor's code
```{r}
election_long <- election_data |>
  pivot_longer(
    cols = -c(Name, State, County),
    names_to = c("metric", "year"),
    names_pattern = "(.+)_(\\d{4})",
    values_to = "pct"
  ) |>
  mutate(year = as.integer(year))
  
  election_tidy <- election_long |>
  pivot_wider(names_from = metric, values_from = pct) |>
  rename(
    county_dem = D,    county_rep = R,
    state_dem  = STD,  state_rep  = STR,
    gap_dem    = DIFD, gap_rep    = DIFR
  ) |>
  mutate(
    county_margin = county_dem - county_rep,
    state_margin  = state_dem  - state_rep,
    margin_gap    = county_margin - state_margin
  )
```

```{r}
lm(margin_gap ~ year, data = election_tidy) 
### coefficients: intercept (-1338.3604, year 0.6734)
### This says that in a linear model of our data, there is an intercept at -1338.3604, and the line we are plotting has a slope of 0.6734. This intercept makes sense given our data begins at x=2000, and it also shows us that college parties have in fact drifted bluer over time. 
```


