---
title: "final_project"
output: html_document
---
CLAUDE LINK: https://claude.ai/share/f888cd21-9563-437c-a19f-d9e91f54c688

###INFO

I found the largest college from every state and looked at the voting results from the past 7 (2000-2024) elections in the counties those colleges are located in. I removed Alaska from the dataset because Alaska does not have counties, and the population is very small and unique. 

This is the source I used to find the largest colleges from each state: https://www.businessinsider.com/map-university-college-most-students-each-state-2014-3

I recorded the proportions of Democratic and Republican popular votes from each election. County votes are under the classifications "D" and "R," and their states are under "STD" and "STR." I also recorded the difference between college counties and the state they are in for democratic and republican proportions, found under classifications DIFD and DIFR. 

In my data, I found evidence of a significant difference in proportion between democratic votes in counties versus their states. 

You can find additional information and previous questions from the original testing in this Google Doc: https://docs.google.com/document/d/1GwR0k3jpDCHCwXB0oGisCfvr7JbEp4jrUfgC5HOg6rg/edit?usp=sharing
 
### Load the tidyverse to read our data
```{r}
library(tidyverse)
```

### Load the data, then we are going to organize the data into simpler datasets.
```{r} 
election_data <- read_csv("CollegeCountyParties.csv")
```
```{r}
election_long <- election_data |>
  pivot_longer(
    cols = -c(Name, State, County),
    names_to = c("metric", "year"),
    names_pattern = "(.+)_(\\d{4})",
    values_to = "pct"
  ) |>
  mutate(year = as.integer(year))
  
  election_tidy <- election_long |>
  pivot_wider(names_from = metric, values_from = pct) |>
  rename(
    county_dem = D,    county_rep = R,
    state_dem  = STD,  state_rep  = STR,
    gap_dem    = DIFD, gap_rep    = DIFR
  ) |>
  mutate(
    county_margin = county_dem - county_rep,
    state_margin  = state_dem  - state_rep,
    margin_gap    = county_margin - state_margin
  )
```
 
   
   
   
### This code makes a linear model using our data to see if the margin has grown

```{r}
model <- lm(margin_gap ~ year, data = election_tidy) 
summary(model)
### coefficients: intercept (-1338.3604, year 0.6734)
### This says that in a linear model of our data, there is an intercept at -1338.3604, and the line we are plotting has a slope of 0.6734. This intercept makes sense given our data begins at x=2000, and it also shows us that college parties have in fact drifted bluer over time. 
### residuals: min -47.092, 1q -11.212, median -0.171, 3q 13.047, max 41.296
### error(year) - 0.1199 t-value 5.616, Pr(>|t|) 4.06e-08
#### residual standard error: 17.77 on 341 degrees of freedom
### multiple r squared: 0.08466 adjusted r squared: 0.08197 
### -> 8.197% of the change in the gap is explained by the time change, so this may not be the best argument to make 
### F-statistic: 31.54 on 1 and df, p-value: 4.062e-08
### while time may not explain the increase, the margin has grown by about # every year. 
```


### Running a significance test to see if there is a significant difference, filtering to the difference in democratic votes, keeping in mind that DIFD = college county minus state percent. 

```{R}
# Filter to only DIFD rows
difd_data <- election_long[election_long$metric == "DIFD", ]
# Run one-sample t-test against mu = 0
t_test_difd <- t.test(
  difd_data$pct,   
  mu = 0,
  alternative = "greater"  # you expect college counties to be MORE Democratic than the state
)

t_test_difd
```
# RESULTS: 

t = 15.897, df = 342, p-value < 2.2e-16
alternative hypothesis: true mean is greater than 0
95 percent confidence interval:
 7.181834      Inf
sample estimates:
mean of x 
 8.013207 
Meaning: We know there is a significant difference in college counties and their states as far as voting democratically. 

### DIFD Mean and SD, finding basic descriptive statistics

```{R}
difd_mean_stats <- difd_data |> 
  summarise(
    mean_pct = mean(pct),
    sd_pct = sd(pct),
    n = n()
  )

difd_mean_stats
#mean = 8.013207
#sd = 9.335507
#n = 343
```
### Creating a simple dataset for the descriptive statistics about the difference in democratic votes. 
```{R}
difd_mean_stats <- difd_mean_stats |>
  mutate(
    se_pct = sd_pct / sqrt(343)
  )

difd_mean_stats
```
### Creating a confidence interval for the true mean of the DIFD

```{R}
difd_mean_stats <- difd_mean_stats |>
  mutate(
    t_critical = qt(0.975, df = n - 1),
    margin_error = t_critical * se_pct,
    ci_lower = mean_pct - margin_error,
    ci_upper = mean_pct + margin_error
  )

difd_mean_stats |>
  select(mean_pct, margin_error, ci_lower, ci_upper)

### mean pct 8.013207
### moe : 0.9914677
### ci lower: 7.021739
### ci upper: 9.004675
### We are 95% confident that the true difference in democratic voting percentage points between a college county and its state is between about 7.02 and 9.00. 
```

### GRAPHS
I wanted to create a graph comparing the average democratic state vote to the average democratic county vote. I asked Claude for help with this as well, so the code I am going to use has been helped by that. There are 2 systems Claude asked to load (dplyr and ggplot2), so that is what we are first loading

```{R}
library(dplyr)
library(ggplot2)

# Filter to just D and STD, then compute averages by year
avg_dem <- election_long %>%
  filter(metric %in% c("D", "STD")) %>%
  group_by(year, metric) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(avg_dem, aes(x = year, y = avg_pct, color = metric)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
    values = c("D" = "blue", "STD" = "lightblue"),
    labels = c("D" = "College County Dem %", "STD" = "State Dem %")
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
  labs(
    title = "Average Democratic Vote: College Counties vs. States",
    x = "Year",
    y = "Average Democratic Vote (%)",
    color = "Group"
  ) +
  theme_minimal()

```
# Comments
This looks like a good graphic to support my hypothesis! 

### Story Pitch: 

#Finding: 
There is data suggesting a significant difference in voting patterns in college counties compared to their states. (Counties containing a large university, in this case the largest university of the state the county belongs to). The data shows that college counties tend to vote about 8 percentage points more Democrat than their states. There is also data suggesting that the gap is growing by about 2 percentage points each election. The data covers the past 7 elections, from 2000 to 2024. 
#Lede:

College counties tend to vote more Democrat

#Potential Issues:
1. While the finding we made was significant and interesting, it's a little unsurprising. Democrats tend to be more educated and younger than Republicans, so when we are looking at counties made up of educated, higher-class, and young individuals, its unsurprising there is a stronger Democratic tendency.
2. I don't want this data to lead readers to any crazy conspiracies. It might be easy to look at this and say "Well big colleges must be brainwashing their students to vote more Democratically." I need to be very careful in my language to not suggest fault. A good way to avoid this is to acknowledge the issues seen in (1). 

#Further information: 



