---
title: "final_project"
output: html_document
---
CLAUDE LINK: https://claude.ai/share/f888cd21-9563-437c-a19f-d9e91f54c688

###INFO

I found the largest college from every state and looked at the voting results from the past 7 (2000-2024) elections in the counties those colleges are located in. I removed Alaska from the dataset because Alaska does not have counties, and the population is very small and unique. 

This is the source I used to find the largest colleges from each state: https://www.businessinsider.com/map-university-college-most-students-each-state-2014-3

I recorded the proportions of Democratic and Republican popular votes from each election. County votes are under the classifications "D" and "R," and their states are under "STD" and "STR." I also recorded the difference between college counties and the state they are in for democratic and republican proportions, found under classifications DIFD and DIFR. 

In my data, I found evidence of a significant difference in proportion between democratic votes in counties versus their states. 

You can find additional information and previous questions from the original testing in this Google Doc: https://docs.google.com/document/d/1GwR0k3jpDCHCwXB0oGisCfvr7JbEp4jrUfgC5HOg6rg/edit?usp=sharing
 
### Load the tidyverse to read our data
```{r}
library(tidyverse)
```

### Load the data, then we are going to organize the data into simpler datasets.
```{r} 
election_data <- read_csv("CollegeCountyParties.csv")
```
```{r}
election_long <- election_data |>
  pivot_longer(
    cols = -c(Name, State, County),
    names_to = c("metric", "year"),
    names_pattern = "(.+)_(\\d{4})",
    values_to = "pct"
  ) |>
  mutate(year = as.integer(year))
  
  election_tidy <- election_long |>
  pivot_wider(names_from = metric, values_from = pct) |>
  rename(
    county_dem = D,    county_rep = R,
    state_dem  = STD,  state_rep  = STR,
    gap_dem    = DIFD, gap_rep    = DIFR
  ) |>
  mutate(
    county_margin = county_dem - county_rep,
    state_margin  = state_dem  - state_rep,
    margin_gap    = county_margin - state_margin
  )
```
#These datasets will make it a little easier to work with our data layer on.
   
   
   
### This code makes a linear model using our data to see if the margin has grown

```{r}
model <- lm(margin_gap ~ year, data = election_tidy) 
summary(model)
### This model is looking at the gap of county_margin (county_dem - county_rep) minus the state margin (state_dem - state_rep). It shows that political divide among counties is increasing compared to the political divide in states.
### The gap between college counties and state voting margins has widened over time
### coefficients: intercept (-1338.3604, year 0.6734)
### This says that in a linear model of our data, there is an intercept at -1338.3604, and the line we are plotting has a slope of 0.6734. This intercept makes sense given our data begins at x=2000, and it also shows us that college parties have in fact drifted farther from their state over time. 
### residuals: min -47.092, 1q -11.212, median -0.171, 3q 13.047, max 41.296
### error(year) - 0.1199 t-value 5.616, Pr(>|t|) 4.06e-08
#### residual standard error: 17.77 on 341 degrees of freedom
### multiple r squared: 0.08466 adjusted r squared: 0.08197 
### -> 8.197% of the change in the gap is explained by the time change, so this may not be the best argument to make 
### F-statistic: 31.54 on 1 and df, p-value: 4.062e-08
### while time may not explain the increase, the margin has grown by about 2.7 every year. 

model <- lm(gap_dem ~ year, data = election_tidy)
summary(model)

### coefficients: intercept (-677.31696, year 0.34062)
### This says that in a linear model of our data, there is an intercept at -677.31696, and the line increase by about 0.34062 every year. Given that an election occurs every 4 years, this means that the gap between college counties and their states' democratic vote percentage points increases by about 1.36248 every election. 
### residuals: min -23.0432, 1q -5.6270, median -0.0958, 3q 6.3831, max 21.518
###error(year) 0.06034, t-value 5.645, Pr(>|t|) 3.49e-08
###residual standard error: 8.941 on 341 degrees of freedom
### multiple r squared 0.08545, adjusted r-squared 0.08277
### -> 8.277% of the change in the gap of democratic votes can be explained by time
### F-statistic, 31.86 on 1 and 341 DF, p-value 3.488e-08
###It would be misleading to just claim that the difference is growing without showing that the value is not the highest, but about 1.4 percentage points of difference every election can be an interesting statistic. 
```


### Running a significance test to see if there is a significant difference, filtering to the difference in democratic votes, keeping in mind that DIFD = college county minus state percent. 

```{R}
# Filter to only DIFD rows
difd_data <- election_long[election_long$metric == "DIFD", ]
# Run one-sample t-test against mu = 0
t_test_difd <- t.test(
  difd_data$pct,   
  mu = 0,
  alternative = "greater"  # you expect college counties to be MORE Democratic than the state
)

t_test_difd
```
# RESULTS: 

t = 15.897, df = 342, p-value < 2.2e-16
alternative hypothesis: true mean is greater than 0
95 percent confidence interval:
 7.181834      Inf
sample estimates:
mean of x 
 8.013207 
Meaning: The analysis provides strong evidence that college counties vote more Democrat than their states on average with a mean in difference that is clearly above 0.  

### DIFD Mean and SD, finding basic descriptive statistics

```{R}
difd_mean_stats <- difd_data |> 
  summarise(
    mean_pct = mean(pct),
    sd_pct = sd(pct),
    n = n()
  )

difd_mean_stats
#mean = 8.013207
#sd = 9.335507
#n = 343
```
### Creating a simple dataset for the descriptive statistics about the difference in democratic votes. 
```{R}
difd_mean_stats <- difd_mean_stats |>
  mutate(
    se_pct = sd_pct / sqrt(343)
  )

difd_mean_stats
```
### Creating a confidence interval for the true mean of the DIFD

```{R}
difd_mean_stats <- difd_mean_stats |>
  mutate(
    t_critical = qt(0.975, df = n - 1),
    margin_error = t_critical * se_pct,
    ci_lower = mean_pct - margin_error,
    ci_upper = mean_pct + margin_error
  )

difd_mean_stats |>
  select(mean_pct, margin_error, ci_lower, ci_upper)

### mean pct 8.013207
### moe : 0.9914677
### ci lower: 7.021739
### ci upper: 9.004675
### We are 95% confident that the true difference in democratic voting percentage points between a college county and its state is between about 7.02 and 9.00. 
```

### GRAPHS
I wanted to create a graph comparing the average democratic state vote to the average democratic county vote. I asked Claude for help with this as well, so the code I am going to use has been helped by that. There are 2 systems Claude asked to load (dplyr and ggplot2), so that is what we are first loading

```{R}
library(dplyr)
library(ggplot2)

# Filter to just D and STD, then compute averages by year
avg_dem <- election_long %>%
  filter(metric %in% c("D", "STD")) %>%
  group_by(year, metric) %>%
  summarise(avg_pct = mean(pct, na.rm = TRUE), .groups = "drop")

# Plot
ggplot(avg_dem, aes(x = year, y = avg_pct, color = metric)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
    values = c("D" = "blue", "STD" = "lightblue"),
    labels = c("D" = "College County Dem %", "STD" = "State Dem %")
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
  labs(
    title = "Average Democratic Vote: College Counties vs. States",
    x = "Year",
    y = "Average Democratic Vote (%)",
    color = "Group"
  ) +
  theme_minimal()

```
# Comments
This looks like a good graphic to support my hypothesis! 

### Story Pitch: 

#Lede:

College counties tend to vote more Democrat

#Finding: 
An analysis of presidential election results from 2000 to 2024 found that counties containing each state's largest university consistently voted more Democratic than their state overall. Across 343 county-election observations, college counties voted on average about 8 percentage points more Democratic than states.

The analysis also suggests that the gap between college counties and state voting margins has widened over time by about 2.7 percentage points each election between a county and its state. A linear regression model found a positive relationship between election year and the difference between county and state voting margins, indicating that college counties have become increasingly more Democratic relative to their belonging state since 2000. Another regression model showed that the democratic percentage point difference is growing by about 1.36 every election. 

#Why it matters: 

Political polarization is often discussed in terms of urban versus rural voters, but college counties may represent another important divide in modern American politics. Universities can influene their local economies, demographics, and culture by concentrating younger voters, highly educated residents, and researchers in one area. This project explores whether those factors are reflected in voting behavior. 

The story would help reinforce the notion that higher educated folks tend to vote more Democratic and help readers understand how higher education may shape regional political ideologies. 

#Evidence to use:
- I'm pretty confident in claiming that the true difference in college counties versus their states is about 8 percentage points. (the p-value was   2.2e-16)
- I am also going to use the linear model's slope (0.6734) multiplied by 4 to show the political gap for elections growing. (2.6936)
- I am going to use the linear model's slope (0.34062) multiplied by 4 (1.36248) to show that data suggests college counties are increasingly voting more blue than their given state.

#Potential Issues:
1. While the finding we made was significant and interesting, it's a little unsurprising. Democrats tend to be more educated and younger than Republicans, so when we are looking at counties made up of educated, higher-class, and young individuals, its unsurprising there is a stronger Democratic tendency.
2. I don't want this data to lead readers to any crazy conspiracies. It might be easy to look at this and say "Well big colleges must be brainwashing their students to vote more Democratically." I need to be very careful in my language to not suggest fault. A good way to avoid this is to acknowledge the issues seen in (1). 
3. Some of these counties have larger populations outside of their college students and include a typical urban area. 
4. The regression model has a small adjusted r-squared value, meaning time alone only explains a little bit of the change in the political divide. This isn't going to be the strongest statistic to report on, which is why I will keep it secondary, but I still think it is interesting. 

#Further information: 
There is some additional information I'd like to collect if I had the chance to. I'd like to examine how much the populations in each of these counties is made up by actual college students, which I could find by examining the ages if that specific information is not readily available. 

As far as interviews to collect, I would interview a political scientist who has an expertise in demographics and ask about voting behavior. I'd also ask about political polarization and contextualizing it in a large college setting. It would be interesting to get some professors from these large universities, and I could also see myself interviewing leaders of political clubs at these schools. It would be interetsing to hear from the Republican students at large universities to see their perspective on the differences. 

As far as further playing with the data, breaking these elections down by year will help strengthen the story and give readers a better timeline of how things have shifted.
